---
title: "PhenoCam Workshop"
output: html_document
date: "2024-04-05"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

------------------------------------------------------------------------

# Purpose

This interactive part of the workshop will walk us through interacting
with the PhenoCam Server using phenocamapi and phenocamr R Package. You
will learn the basics of extracting PhenoCam data and metadata and how
to work with that data through these packages. This code was adopted
from code by Bijan Seyednasrollah and Koen Hufkens.

------------------------------------------------------------------------

We will begin with several useful skills and tools for extracting
PhenoCam data directly from the server. The goals of this portion of the
workshop are to:

-   Explore the PhenoCam metadata
-   Download PhenoCam time-series data
-   Download midday images for a given time range
-   Extract and plot the phenophases from a site of interest.

## Exploring PhenoCam metadata

Each PhenoCam site has specific metadata including but not limited to
how a site is set up and where it is located, what vegetation type is
visible from the camera, and its climate regime.

Each PhenoCam may have zero to several Regions of Interest (ROIs) per
vegetation type. The packages that we will use today allow us to
interact with the PhenoCam server to extract PhenoCam data and process
that data in an R environment.

Let's bring in the packages that we will be using!

```{r libraries}
library(data.table) #installs package that creates a data frame for visualizing data in row-column table format
library(phenocamapi)  #installs packages of time series and phenocam data from the Phenology Network. Loads required packages rjson, bitops and RCurl
library(lubridate)  #install time series data package
library(jpeg)
library(phenocamr)
```

We can obtain an up-to-date `data.frame` of the metadata of the entire
PhenoCam network using the `get_phenos()` function. The returning value
would be a `data.table` in order to simplify further data exploration.

```{r obtain-data, fig.height=5, fig.width=8, message=FALSE}

#Obtain phenocam metadata from the Phenology Network in form of a data.table
phenos <- get_phenos()

#Explore metadata table
head(phenos$site) #preview first six rows of the table. These are the first six phenocam sites in the Phenology Network

colnames(phenos)  #view all column names. 
#This is all the metadata we have for the phenocams in the Phenology Network
```

Now we have a better idea of the types of metadata that are available
for the Phenocams.

### Remove null values

We may want to explore some of the patterns in the metadata before we
jump into specific locations. Let's look at Mean Annual Precipitation
(MAP) and Mean Annual Temperature (MAT) across the different field site
and classify those by the primary vegetation type ('primary_veg_type')
for each site. First we need to identify the abbreviations that are used
for the functions that we will be using next. We will use the table
below to do this.

| Abbreviation |                     Description                     |     |
|--------------|:---------------------------------------------------:|----:|
| AG           |                     agriculture                     |     |
| DB           |                 deciduous broadleaf                 |     |
| DN           |                deciduous needleleaf                 |     |
| EB           |                 evergreen broadleaf                 |     |
| EN           |                evergreen needleleaf                 |     |
| GR           |                      grassland                      |     |
| MX           | mixed vegetation (generally EN/DN, DB/EN, or DB/EB) |     |
| SH           |                       shrubs                        |     |
| TN           |   tundra (includes sedges, lichens, mosses, etc.)   |     |
| WT           |                       wetland                       |     |
| NV           |                    non-vegetated                    |     |
| RF           |                   reference panel                   |     |
| XX           |                     unspecified                     |     |

First, we want to remove the sites where there is not MAP and MET as
part of the metadata, and then plot the data.

```{r plot-MAT-MAP, message=FALSE, fig.height=8, fig.width=8}
# #Some sites do not have data on Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT).

# removing the sites with unknown MAT and MAP values
phenos <- phenos[!((MAT_worldclim == -9999)|(MAP_worldclim == -9999))]

# Making a plot showing all sites by their vegetation type (represented as different symbols and colors) plotting across climate (MAT and MAP) space. Refer to table to identify vegetation type acronyms.
phenos[primary_veg_type=='DB', plot(MAT_worldclim, MAP_worldclim, pch = 19, col = 'green', xlim = c(-5, 27), ylim = c(0, 4000))]
phenos[primary_veg_type=='DN', points(MAT_worldclim, MAP_worldclim, pch = 1, col = 'darkgreen')]
phenos[primary_veg_type=='EN', points(MAT_worldclim, MAP_worldclim, pch = 17, col = 'brown')]
phenos[primary_veg_type=='EB', points(MAT_worldclim, MAP_worldclim, pch = 25, col = 'orange')]
phenos[primary_veg_type=='AG', points(MAT_worldclim, MAP_worldclim, pch = 12, col = 'yellow')]
phenos[primary_veg_type=='SH', points(MAT_worldclim, MAP_worldclim, pch = 23, col = 'red')]

legend('topleft', legend = c('DB','DN', 'EN','EB','AG', 'SH'), 
       pch = c(19, 1, 17, 25, 12, 23), 
       col =  c('green', 'darkgreen', 'brown',  'orange',  'yellow',  'red' ))

```

## PhenoCam time series

PhenoCam time series are extracted time series data obtained from
regions of interest (ROI's) for a given site. First we have to obtain
the ROIs.

### Obtain ROIs

To download the phenological time series from the PhenoCam, we need to
know the site name, vegetation type and ROI ID. This information can be
obtained from each specific PhenoCam page by using the `get_rois()`
function from the phenocamapi package.

```{r get-rois, fig.height=5, fig.width=6.5, message=FALSE}
# Obtaining the list of all the available regions of interest (ROI's) on the PhenoCam server and producing a data table
rois <- get_rois()

# view the data variables in the data table
colnames(rois)

# view first few regions of of interest (ROI) locations
head(rois$roi_name)
```

### Download time series

Now that we have our ROI data frame, we can use the `get_pheno_ts()`
function to download a time series of our site within that region of
interest and return the result as a `data.table`. We will work with the
Duke Forest Hardwood Stand ('dukehw') and the ROI == 1000.

```{r, fig.height=5, fig.width=6.5, message=FALSE}
# list ROIs for dukehw
rois[site=='dukehw',]

# Obtain the decidous broadleaf, ROI ID 1000 data from the dukehw phenocam
dukehw_DB_1000 <- get_pheno_ts(site = 'dukehw', vegType = 'DB', roiID = 1000, type = '3day')

# Produces a list of the dukehw data variables
str(dukehw_DB_1000)

```

We now have a variety of data related to this ROI from the Hardwood
Stand at Duke Forest. One of those is the green chromatic coordinate.
This is a measure of "greenness" of an area and is an indicator of the
green pigment in vegetation.

We can use this to measure changes in GCC or greeness over time at this
site. We will use what's called `gcc90`, which is the the 90th quantile
of GCC in the pixels across the ROI. The 90th quantile tracks the upper
greenness values while not including many outliers.

Before we can plot `gcc-90` we do need to fix our dates and convert them
from Factors to Date to correctly plot.

```{r plot-gcc90, fig.height=5, fig.width=8}
# Convert date variable into date format
dukehw_DB_1000[,date:=as.Date(date)]

# plot gcc_90
dukehw_DB_1000[,plot(date, gcc_90, col = 'green', type = 'b')]
mtext('Duke Forest, Hardwood', font = 2)

```

### Download midday images for a given time range

But now let's say we want to actually "peak" into our forest. To do this
we can access all the midday images and download them one at a time.
However, we frequently want all the images within a specific time range
of interest. We can do that using the the phenocamapi package.

```{r midday-time-range, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}

# open a temporary directory
tmp_dir <- tempdir()

# download a subset. Example dukehw 2017
download_midday_images(site = 'dukehw', # which site
                       y = 2017, # which year(s)
                       months = 1:12, # which month(s)
                       days = 15, # which days on month(s)
                       download_dir = tmp_dir) # where on your computer

# list of downloaded files
duke_middays_path <- dir(tmp_dir, pattern = 'dukehw*', full.names = TRUE)

head(duke_middays_path)

```

And with these images, we can demonstrate the seasonality of Duke forest
observed from the camera. (Note this code may take a while to run
through the loop).

```{r plot-monthly-forest, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}
n <- length(duke_middays_path)
par(mar= c(0,0,0,0), mfrow=c(4,3), oma=c(0,0,3,0))

for(i in 1:n){
  img <- readJPEG(duke_middays_path[i])
  plot(0:1,0:1, type='n', axes= FALSE, xlab= '', ylab = '')
  rasterImage(img, 0, 0, 1, 1)
  mtext(month.name[i], line = -2)
}
mtext('Seasonal variation of forest at Duke Hardwood Forest', font = 2, outer = TRUE)

```

# Phenology Transition Dates

Okay, great! We just visualized an entire growing season through
PhenoCam images. But now what if we want to see if this growing season
has been changing through time? Or track if start of sesaon, end of
season, or greeness has changed through time? We can use the phenocamr
package to visualize this. The phenphases() function in the phenocamr
package will extract the start of spring and end of fall dates. First
let's download the duke forest data using this package.

```{r}
 download_phenocam(site = "dukehw",
                    veg_type = "DB",
                    roi_id = "1000",
                    frequency = 3,
                    outlier_detection = FALSE,
                    smooth = FALSE,
                    out_dir = tempdir())

#Now we need to assign a dataframe to the data that we just downloaded
df <- read_phenocam(file.path(tempdir(),"dukehw_DB_1000_3day.csv"))
```

# Detecting Outliers and Smoothing the Time Series

Now that we have our dataframe, we will want to detect any outliers in
our data and smooth the time series.

```{r}
  df <- detect_outliers(df)
  df <- smooth_ts(df)
```

# Calculating the Phenophases

Now that the smoothed data has been made available, we can calculate and
plot the phenological transition date.

```{r}
phenology_dates <- phenophases(df, internal = TRUE)

plot(as.Date(df$data$date),
     df$data$smooth_gcc_90,
     type = "l",
     xlab = "date",
     ylab = "Gcc")

# rising "spring" greenup dates
abline(v = phenology_dates$rising$transition_50,
       col = "forestgreen")

# falling "autumn" senescence dates
abline(v = phenology_dates$falling$transition_50,
       col = "brown4")
```

------------------------------------------------------------------------
