---
title: "PhenoCam Workshop"
output: html_document
date: "2024-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# Purpose 
This interactive part of the workshop will walk participants through interacting with the PhenoCam Server using phenocamapi and phenocamr R Package. Participants will learn the basics of extracting PhenoCam data and metadata and how to work with that data through these packages. This code was adopted from code by Bijan Seyednasrollah and Koen Hufkens.  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


We will begin with several useful skills and tools for extracting PhenoCam data 
directly from the server. The goals of this portion of the workshop are to:

- Explore the PhenoCam metadata
- Filter the dataset by site attributes
- Download PhenoCam time-series data
- Extract the list of midday images
- Download midday images for a given time range
- Extract and plot the phenophases from a site of interest. 

## Exploring PhenoCam metadata

Each PhenoCam site has specific metadata including but not limited to how a site 
is set up and where it is located, what vegetation type is visible from the 
camera, and its climate regime. Each PhenoCam may have zero to several Regions 
of Interest (ROIs) per vegetation type. The *phenocamapi* package is an 
interface to interact with the PhenoCam server to extract those data and 
process them in an R environment.

To explore the PhenoCam data, we'll use several packages for this tutorial. 

```{r libraries}
if(!require(devtools)) install.packages('devtools')
devtools::install_github('bnasr/phenocamapi')
library(data.table) #installs package that creates a data frame for visualizing data in row-column table format
library(phenocamapi)  #installs packages of time series and phenocam data from the Phenology Network. Loads required packages rjson, bitops and RCurl
library(lubridate)  #install time series data package
library(jpeg)
```


We can obtain an up-to-date `data.frame` of the metadata of the entire PhenoCam 
network using the `get_phenos()` function. The returning value would be a 
`data.table` in order to simplify further data exploration.

```{r obtain-data, fig.height=5, fig.width=8, message=FALSE}

#Obtain phenocam metadata from the Phenology Network in form of a data.table
phenos <- get_phenos()

#Explore metadata table
head(phenos$site) #preview first six rows of the table. These are the first six phenocam sites in the Phenology Network

colnames(phenos)  #view all column names. 
#This is all the metadata we have for the phenocams in the Phenology Network
```

Now we have a better idea of the types of metadata that are available for the
Phenocams. 

### Remove null values

We may want to explore some of the patterns in the metadata before we jump into
specific locations. Let's look at Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) across the different field site and classify those by the primary vegetation type ('primary_veg_type') for each site. 

| Abbreviation |	Description |
|----------|:-------------:|------:|
| AG |	agriculture |
| DB |	deciduous broadleaf |
| DN |	deciduous needleleaf |
| EB |	evergreen broadleaf |
| EN |	evergreen needleleaf |
| GR |	grassland |
| MX |	mixed vegetation (generally EN/DN, DB/EN, or DB/EB) |
| SH |	shrubs |
| TN |	tundra (includes sedges, lichens, mosses, etc.) |
| WT |	wetland |
| NV |	non-vegetated |
| RF |	reference panel | 	 
| XX |	unspecified |

To do this we'd first want to remove the sites where there is not data and then 
plot the data. 

```{r plot-MAT-MAP, message=FALSE, fig.height=8, fig.width=8}
# #Some sites do not have data on Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT).

# removing the sites with unknown MAT and MAP values
phenos <- phenos[!((MAT_worldclim == -9999)|(MAP_worldclim == -9999))]

# Making a plot showing all sites by their vegetation type (represented as different symbols and colors) plotting across climate (MAT and MAP) space. Refer to table to identify vegetation type acronyms.
phenos[primary_veg_type=='DB', plot(MAT_worldclim, MAP_worldclim, pch = 19, col = 'green', xlim = c(-5, 27), ylim = c(0, 4000))]
phenos[primary_veg_type=='DN', points(MAT_worldclim, MAP_worldclim, pch = 1, col = 'darkgreen')]
phenos[primary_veg_type=='EN', points(MAT_worldclim, MAP_worldclim, pch = 17, col = 'brown')]
phenos[primary_veg_type=='EB', points(MAT_worldclim, MAP_worldclim, pch = 25, col = 'orange')]
phenos[primary_veg_type=='AG', points(MAT_worldclim, MAP_worldclim, pch = 12, col = 'yellow')]
phenos[primary_veg_type=='SH', points(MAT_worldclim, MAP_worldclim, pch = 23, col = 'red')]

legend('topleft', legend = c('DB','DN', 'EN','EB','AG', 'SH'), 
       pch = c(19, 1, 17, 25, 12, 23), 
       col =  c('green', 'darkgreen', 'brown',  'orange',  'yellow',  'red' ))

```



## PhenoCam time series

PhenoCam time series are extracted time series data obtained from regions of interest (ROI's) for a given site. First we have to obtain the ROIs.

### Obtain ROIs
To download the phenological time series from the PhenoCam, we need to know the 
site name, vegetation type and ROI ID. This information can be obtained from each
specific PhenoCam page on the 
<a href="https://phenocam.sr.unh.edu/webcam/gallery/" target="_blank">PhenoCam website</a> 
or by using the `get_rois()` function. 


```{r get-rois, fig.height=5, fig.width=6.5, message=FALSE}
# Obtaining the list of all the available regions of interest (ROI's) on the PhenoCam server and producing a data table
rois <- get_rois()

# view the data variables in the data table
colnames(rois)

# view first few regions of of interest (ROI) locations
head(rois$roi_name)
```

### Download time series

The `get_pheno_ts()` function can download a time series and return the result 
as a `data.table`. 
Let's work with the 
<a href="https://phenocam.sr.unh.edu/data/archive/dukehw/ROI/dukehw_DB_1000.html">Duke Forest Hardwood Stand (`dukehw`) PhenoCam</a> 
and specifically the ROI 
<a href="https://phenocam.sr.unh.edu/data/archive/dukehw/ROI/dukehw_DB_1000.html">`DB_1000`</a> 
we can run the following code. 

```{r, fig.height=5, fig.width=6.5, message=FALSE}
# list ROIs for dukehw
rois[site=='dukehw',]

# Obtain the decidous broadleaf, ROI ID 1000 data from the dukehw phenocam
dukehw_DB_1000 <- get_pheno_ts(site = 'dukehw', vegType = 'DB', roiID = 1000, type = '3day')

# Produces a list of the dukehw data variables
str(dukehw_DB_1000)

```

We now have a variety of data related to this ROI from the Hardwood Stand at Duke 
Forest. 

Green Chromatic Coordinate (GCC) is a measure of "greenness" of an area and is
widely used in Phenocam images as an indicator of the green pigment in vegetation.
Let's use this measure to look at changes in GCC over time at this site. Looking 
back at the available data, we have several options for GCC. `gcc90` is the 90th 
quantile of GCC in the pixels across the ROI (for more details, 
<a href="https://daac.ornl.gov/VEGETATION/guides/PhenoCam_V1.html" target="_blank"> PhenoCam v1 description</a>). 
We'll use this as it tracks the upper greenness values while not including many 
outliers.  

Before we can plot `gcc-90` we do need to fix our dates and convert them from 
Factors to Date to correctly plot. 

```{r plot-gcc90, fig.height=5, fig.width=8}
# Convert date variable into date format
dukehw_DB_1000[,date:=as.Date(date)]

# plot gcc_90
dukehw_DB_1000[,plot(date, gcc_90, col = 'green', type = 'b')]
mtext('Duke Forest, Hardwood', font = 2)

```


### Download midday images for a given time range

Now we can access all the midday images and download them one at a time. However,
we frequently want all the images within a specific time range of interest. We'll
learn how to do that next. 

```{r midday-time-range, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}

# open a temporary directory
tmp_dir <- tempdir()

# download a subset. Example dukehw 2017
download_midday_images(site = 'dukehw', # which site
                       y = 2017, # which year(s)
                       months = 1:12, # which month(s)
                       days = 15, # which days on month(s)
                       download_dir = tmp_dir) # where on your computer

# list of downloaded files
duke_middays_path <- dir(tmp_dir, pattern = 'dukehw*', full.names = TRUE)

head(duke_middays_path)

```

We can demonstrate the seasonality of Duke forest observed from the camera. (Note
this code may take a while to run through the loop).
```{r plot-monthly-forest, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}
n <- length(duke_middays_path)
par(mar= c(0,0,0,0), mfrow=c(4,3), oma=c(0,0,3,0))

for(i in 1:n){
  img <- readJPEG(duke_middays_path[i])
  plot(0:1,0:1, type='n', axes= FALSE, xlab= '', ylab = '')
  rasterImage(img, 0, 0, 1, 1)
  mtext(month.name[i], line = -2)
}
mtext('Seasonal variation of forest at Duke Hardwood Forest', font = 2, outer = TRUE)

```

# Phenology Transition Dates
The goal of this section was to show how to download a limited number of midday images from the PhenoCam server. However, more extensive datasets should be downloaded from the <a href="https://phenocam.sr.unh.edu/webcam/network/download/"> PhenoCam </a>.

Now what if we want to see rising and falling of the seasons on a plot through time? This let's us visualize how the phenology might have changed through time in an ecosystem. To do this, we can use the phenocamr package. The phenphases() function in the phenocamr package, but first let's download the duke forest data using this package.

```{r}
 download_phenocam(site = "dukehw",
                    veg_type = "DB",
                    roi_id = "1000",
                    frequency = 3,
                    outlier_detection = FALSE,
                    smooth = FALSE,
                    out_dir = tempdir())

#Now we need to assign a dataframe to the data that we just downloaded
df <- read_phenocam(file.path(tempdir(),"dukehw_DB_1000_3day.csv"))
```


# Detecting Outliers and Smoothing the Time Series

Now that we have our dataframe, we will want to detect any outliers in our data and smooth the time series 
```{r}
  df <- detect_outliers(df)
  df <- smooth_ts(df)
```

# Calculating the Phenophases
Now that the smoothed data has been made available, we can calculate and plot the phenological transition date.
```{r}
phenology_dates <- phenophases(df, internal = TRUE)

plot(as.Date(df$data$date),
     df$data$smooth_gcc_90,
     type = "l",
     xlab = "date",
     ylab = "Gcc")

# rising "spring" greenup dates
abline(v = phenology_dates$rising$transition_50,
       col = "forestgreen")

# falling "autumn" senescence dates
abline(v = phenology_dates$falling$transition_50,
       col = "brown4")
```



*** 

