---
---
title: "Workshop Exercise"
output: html_document
date: "2024-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Exploring PhenoCam metadata

In this part of the workshop, the participant will select their own site to explore. First, we need to install and load the approporiate libraries.

```{r libraries}
if(!require(devtools)) install.packages('devtools')
devtools::install_github('bnasr/phenocamapi')
library(data.table) #installs package that creates a data frame for visualizing data in row-column table format
library(phenocamapi)  #installs packages of time series and phenocam data from the Phenology Network. Loads required packages rjson, bitops and RCurl
library(lubridate)  #install time series data package
library(jpeg)
```


#Download the PhenoCam data

Now we will download the metadata of the entire PhenoCam network using the 'get_phenos' function. This will return a 'data.table' to continue exploring the data.  

```{r obtain-data, fig.height=5, fig.width=8, message=FALSE}

#Obtain phenocam metadata from the Phenology Network in form of a data.table
mydf <- get_phenos()

#Explore metadata table
head(mydf$site) #preview first six rows of the table. These are the first six phenocam sites in the Phenology Network

colnames(mydf)  #view all column names. 
#This is all the metadata we have for the phenocams in the Phenology Network
```

Now we have a better idea of the types of metadata that are available for the
Phenocams. 

### Remove null values

At this point, select a few vegetation types that you might be interested in exploring and plot their Mean Annual Precipitation and Mean Annual Temperature. See the table below for the vegetation abbreviation. To do this we'd first want to remove the sites where there is not data and then 
plot the data. 

| Abbreviation |	Description |
|----------|:-------------:|------:|
| AG |	agriculture |
| DB |	deciduous broadleaf |
| DN |	deciduous needleleaf |
| EB |	evergreen broadleaf |
| EN |	evergreen needleleaf |
| GR |	grassland |
| MX |	mixed vegetation (generally EN/DN, DB/EN, or DB/EB) |
| SH |	shrubs |
| TN |	tundra (includes sedges, lichens, mosses, etc.) |
| WT |	wetland |
| NV |	non-vegetated |
| RF |	reference panel | 	 
| XX |	unspecified |


```{r plot-MAT-MAP, message=FALSE, fig.height=8, fig.width=8}
# #Some sites do not have data on Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT).

# removing the sites with unknown MAT and MAP values
mydf <- mydf[!((MAT_worldclim == -9999)|(MAP_worldclim == -9999))]

# Making a plot showing all sites by their vegetation type (represented as different symbols and colors) plotting across climate (MAT and MAP) space. Refer to table to identify vegetation type acronyms.
mydf[primary_veg_type=='insertvegtypeherefromtable', plot(MAT_worldclim, MAP_worldclim, pch = 19, col = 'green', xlim = c(-5, 27), ylim = c(0, 4000))]
mydf[primary_veg_type=='insertvegtypeherefromtable', points(MAT_worldclim, MAP_worldclim, pch = 1, col = 'darkgreen')]
mydf[primary_veg_type=='insertvegtypeherefromtable', points(MAT_worldclim, MAP_worldclim, pch = 17, col = 'brown')]

legend('topleft', legend = c('DB','DN', 'EN'), 
       pch = c(19, 1, 17), 
       col =  c('green', 'darkgreen', 'brown'))

```


### Filtering using attributes

Now you can filter your dataframe by certain attributes in other datasets, such as flux data. Let's do this for your dataframe using the `flux_sitenames`! 


```{r filter-flux, fig.height=5, fig.width=6.5, message=FALSE}
# Create a data table only including the sites that have flux_data available and where the FLUX site name is specified
myphenofluxsites <- mydf[flux_data==TRUE&!is.na(flux_sitenames)&flux_sitenames!='', 
                         .(PhenoCam=site, Flux=flux_sitenames)] # return as table

#Specify to retain variables of Phenocam site and their flux tower name
myphenofluxsites <- myphenofluxsites[Flux!='']

# view the first few rows of the data table
head(myphenofluxsites)

```

Now select one single vegetation type that you are interested in exploring. For example, you could select a flux tower that is located in deciduous broadleaf forests (`primary_veg_type=='DB'`). Use the table above to select the vegetation type you want to work with.

```{r filter-flux-db, fig.height=5, fig.width=6.5, message=FALSE}

#list deciduous broadleaf sites with a flux tower
my.flux <- mydf[flux_data==TRUE&primary_veg_type=='InsertVegetationTypeFromTable', 
                  site]  # return just the site names as a list

# see the first few rows
head(my.flux)
```


## PhenoCam time series

PhenoCam time series are extracted time series data obtained from regions of interest (ROI's) for a given site. 

### Obtain ROIs
To download the phenological time series from the PhenoCam, we need to know the 
site name, vegetation type and ROI ID. This information can be obtained from each
specific PhenoCam page on the 
<a href="https://phenocam.sr.unh.edu/webcam/gallery/" target="_blank">PhenoCam website</a> 
or by using the `get_rois()` function. 


```{r get-rois, fig.height=5, fig.width=6.5, message=FALSE}
# Obtaining the list of all the available regions of interest (ROI's) on the PhenoCam server and producing a data table
myrois <- get_rois()

# view the data variables in the data table
colnames(myrois)

# view first few regions of of interest (ROI) locations
head(myrois$roi_name)

#Explore the list of sites and choose a site from the list of sites that prints out below. 
myrois$site

#Prints out the vegetation type for the site that you select, where the row of your site corresponds to the row of the vegetation type.
rois$roitype
```

### Download time series

The `get_pheno_ts()` function can download a time series and return the result 
as a `data.table`. 
Let's work with the 
<a href="https://phenocam.sr.unh.edu/data/archive/dukehw/ROI/dukehw_DB_1000.html">Duke Forest Hardwood Stand (`dukehw`) PhenoCam</a> 
and specifically the ROI 
<a href="https://phenocam.sr.unh.edu/data/archive/dukehw/ROI/dukehw_DB_1000.html">`DB_1000`</a> 
we can run the following code. Make sure that you insert the site name that you selected from the code chunk above, replacing the 'insertsitename.' 

```{r, fig.height=5, fig.width=6.5, message=FALSE}
# list ROIs for dukehw
rois[site=='insertsitename',]

# Obtain the decidous broadleaf, ROI ID 1000 data from the dukehw phenocam
mysite_DB_1000 <- get_pheno_ts(site = 'insertsitename', vegType = 'insertvegetationtype', roiID = 1000, type = '3day')

# Produces a list of the dukehw data variables
str(mysite_DB_1000)

```

We now have a variety of data related to this ROI from the Hardwood Stand at Duke 
Forest. 

Green Chromatic Coordinate (GCC) is a measure of "greenness" of an area and is
widely used in Phenocam images as an indicator of the green pigment in vegetation.
Let's use this measure to look at changes in GCC over time at this site. Looking 
back at the available data, we have several options for GCC. `gcc90` is the 90th 
quantile of GCC in the pixels across the ROI (for more details, 
<a href="https://daac.ornl.gov/VEGETATION/guides/PhenoCam_V1.html" target="_blank"> PhenoCam v1 description</a>). 
We'll use this as it tracks the upper greenness values while not including many 
outliners.  

Before we can plot `gcc-90` we do need to fix our dates and convert them from 
Factors to Date to correctly plot. 

```{r plot-gcc90, fig.height=5, fig.width=8}
# Convert date variable into date format
mysite_DB_1000[,date:=as.Date(date)]

# plot gcc_90
mysite_DB_1000[,plot(date, gcc_90, col = 'green', type = 'b')]
mtext('My Site GCC', font = 2)

```


## Download midday images

While PhenoCam sites may have many images in a given day, many simple analyses
can use just the midday image when the sun is most directly overhead the canopy. 
Therefore, extracting a list of midday images (only one image a day) can be useful. Let's do this for the site that you selected! Replace 'insertsitename' with the name of your site. 


```{r midday-list, fig.height=5, fig.width=8, message=FALSE}

# obtaining midday_images for dukehw
mysite_middays <- get_midday_list('insertsitename')

# see the first few rows
head(mysite_middays)

```

Now we have a list of all the midday images from this Phenocam. Let's download
them and plot them for your site. 

```{r midday-download, fig.height=5, fig.width=8}
# download a file
destfile <- tempfile(fileext = '.jpg')

# download only the first available file
# modify the `[1]` to download other images
download.file(mysite_middays[1], destfile = destfile, mode = 'wb')

# plot the image
img <- try(readJPEG(destfile))
if(class(img)!='try-error'){
  par(mar= c(0,0,0,0))
  plot(0:1,0:1, type='n', axes= FALSE, xlab= '', ylab = '')
  rasterImage(img, 0, 0, 1, 1)
}
```


### Download midday images for a given time range

Now we can access all the midday images and download them one at a time. However,
we frequently want all the images within a specific time range of interest. We'll
learn how to do that next. Replace 'mysitename' with the name of the site you selected.  

```{r midday-time-range, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}

# open a temporary directory
tmp_dir <- tempdir()

# download a subset. Example dukehw 2017
download_midday_images(site = 'mysitename', # which site
                       y = 2017, # which year(s)
                       months = 1:12, # which month(s)
                       days = 15, # which days on month(s)
                       download_dir = tmp_dir) # where on your computer

# list of downloaded files
mysite_middays_path <- dir(tmp_dir, pattern = 'mysitename*', full.names = TRUE)

head(mysite_middays_path)

```

We can demonstrate the seasonality of your vegetation type observed from the camera. (Note
this code may take a while to run through the loop).
```{r plot-monthly-forest, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}
n <- length(mysite_middays_path)
par(mar= c(0,0,0,0), mfrow=c(4,3), oma=c(0,0,3,0))

for(i in 1:n){
  img <- readJPEG(mysite_middays_path[i])
  plot(0:1,0:1, type='n', axes= FALSE, xlab= '', ylab = '')
  rasterImage(img, 0, 0, 1, 1)
  mtext(month.name[i], line = -2)
}
mtext('Seasonal variation of my vegetation type and site', font = 2, outer = TRUE)

```

# Phenology Transition Dates
The goal of this section was to show how to download a limited number of midday images from the PhenoCam server. However, more extensive datasets should be downloaded from the <a href="https://phenocam.sr.unh.edu/webcam/network/download/"> PhenoCam </a>.

Now what if we want to see rising and falling of the seasons on a plot through time? This let's us visualize how the phenology might have changed through time in an ecosystem. To do this, we can use the phenocamr package. The phenphases() function in the phenocamr package, but first let's download the duke forest data using this package. Replace "mysitename" with the name of your site that you selected.

```{r}
 download_phenocam(site = "mysitename",
                    veg_type = "DB",
                    roi_id = "1000",
                    frequency = 3,
                    outlier_detection = FALSE,
                    smooth = FALSE,
                    out_dir = tempdir())

#Now we need to assign a dataframe to the data that we just downloaded
dfphenophases <- read_phenocam(file.path(tempdir(),"mysitename_DB_1000_3day.csv"))
```


# Detecting Outliers and Smoothing the Time Series

Now that we have our dataframe, we will want to detect any outliers in our data and smooth the time series from your site. 
```{r}
  dfphenophases <- detect_outliers(dfphenophases)
  dfphenophases <- smooth_ts(dfphenophases)
```

# Calculating the Phenophases
Now that the smoothed data has been made available, we can calculate and plot the phenological transition date for your site.
```{r}
myphenology_dates <- phenophases(dfphenophases, internal = TRUE)

plot(as.Date(dfphenophases$data$date),
     dfphenophases$data$smooth_gcc_90,
     type = "l",
     xlab = "date",
     ylab = "Gcc")

# rising "spring" greenup dates
abline(v = myphenology_dates$rising$transition_50,
       col = "forestgreen")

# falling "autumn" senescence dates
abline(v = myphenology_dates$falling$transition_50,
       col = "brown4")
```



*** 
