---
title: "Workshop Exercise"
output: html_document
date: "2024-04-05"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Start Here

## Exploring PhenoCam metadata

In this part of the workshop, you will select your own site to explore.
First, we need to install and load the appropriate libraries.
[Run]{style="color:green"} the chunk of code below by selecting the
green arrow in the code chunk.

```{r libraries}
library(data.table) #installs package that creates a data frame for visualizing data in row-column table format
library(phenocamapi)  #installs packages of time series and phenocam data from the Phenology Network. Loads required packages rjson, bitops and RCurl
library(lubridate)  #install time series data package
library(jpeg)
library(phenocamr)
library(dplyr)
```

Now we will download the metadata of the entire PhenoCam network using
the 'get_phenos' function. This will return a 'data.table' to continue
exploring the data. Run the code below to do this. We will call the
data.table that is returned from this function 'mydf'.

Explore the column names that are printed out from this code chunk. The
column names indicate the metadata that is downloaded from the PhenoCam
website. [Run]{style="color:green"} the code chunk below without any
changes to the code to explore the output.

```{r obtain-data, fig.height=5, fig.width=8, message=FALSE}
#Obtain phenocam metadata from the Phenology Network in form of a data.table. This data.table will be called 'mydf'
mydf <- get_phenos()

#Explore metadata table
head(mydf$site) #preview first six rows of the table. These are the first six phenocam sites in the Phenology Network

colnames(mydf)  #view all column names that are downloaded as part of the phenocam metadata. 
#This is all the metadata we have for the phenocams in the Phenology Network
```

Now we have a better idea of the types of metadata that are available
from the Phenocams.

### Remove null values and explore metadata across vegetation types.

| Abbreviation |                     Description                     |     |
|--------------|:---------------------------------------------------:|----:|
| AG           |                     agriculture                     |     |
| DB           |                 deciduous broadleaf                 |     |
| DN           |                deciduous needleleaf                 |     |
| EB           |                 evergreen broadleaf                 |     |
| EN           |                evergreen needleleaf                 |     |
| GR           |                      grassland                      |     |
| MX           | mixed vegetation (generally EN/DN, DB/EN, or DB/EB) |     |
| SH           |                       shrubs                        |     |
| TN           |   tundra (includes sedges, lichens, mosses, etc.)   |     |
| WT           |                       wetland                       |     |
| NV           |                    non-vegetated                    |     |
| RF           |                   reference panel                   |     |
| XX           |                     unspecified                     |     |

At this point, [choose two vegetation types]{style="color:green"} from
the table above that you might be interested in exploring to plot their
Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAP).

[See the table above to select two vegetation abbreviations of
interest.]{style="color:green"}.

The function below will accept the vegetation abbreviation rather than
the full name or description. To do this we will first want to remove
the sites where there is no MAT or MAP data available and then plot the
data from the vegetation types you select.

For this, [we will work with the data.table called 'my.df' from
above.]{style="color:green"} [Replace
'InsertVegetationAbbreviationFromTable' with the vegetation type you
selected.]{style="color:red" Once this is complete,
[run]{style="color:green"} the code chunk below.

[Example:]{style="color:red"} let's select DB and DN:
mydf[primary_veg_type=='DB', plot(MAT_worldclim, MAP_worldclim, pch =
19, col = 'green', xlim = c(-5, 27), ylim = c(0, 4000))]
mydf[primary_veg_type=='DN', points(MAT_worldclim, MAP_worldclim, pch =
1, col = 'darkgreen')]

legend('topleft', legend = c('DB','DN'), pch = c(19, 1), col =
c('green', 'darkgreen'))

```{r plot-MAT-MAP, message=FALSE, fig.height=8, fig.width=8}
# #Some sites do not have data on Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT), so we are removing data where there is no MAP or MAT.

# removing the sites with unknown MAT and MAP values
mydf <- mydf[!((MAT_worldclim == -9999)|(MAP_worldclim == -9999))]

# Making a plot showing all sites by their vegetation type (represented as different symbols and colors) plotting across climate (MAT and MAP) space. Refer to table to identify vegetation type acronyms.

mydf[primary_veg_type=='InsertVegetationAbbreviationFromTable', plot(MAT_worldclim, MAP_worldclim, pch = 19, col = 'green', xlim = c(-5, 27), ylim = c(0, 4000))]
mydf[primary_veg_type=='InsertVegetationAbbreviationFromTable', points(MAT_worldclim, MAP_worldclim, pch = 1, col = 'darkgreen')]


legend('topleft', legend = c('InsertVegetationAbbreviationFromTable','InsertVegetationAbbreviationFromTable'), 
       pch = c(19, 1), 
       col =  c('green', 'darkgreen'))

```

## PhenoCam time series

PhenoCam time series are extracted time series data obtained from
regions of interest (ROI's) for a given site.

### Obtain ROIs

To download the phenological time series from the PhenoCam network, we
need to know the site name, vegetation type and ROI ID. This information
can be obtained from each specific PhenoCam site by using the
`get_rois()` function. This function connects directly to the PhenoCam
website. For the purposes of this exercise, we will use the roi
dataframe that we have already downloaded.

The data.table that is returned is called 'myrois'. Let explore what
data is downloaded using this function by exploring the data.table.
[Run]{style="color:green"} the code chunk below to do this using the
green arrow in the code chunk without any changes to the code.

```{r get-rois, fig.height=5, fig.width=6.5, message=FALSE}
# Obtaining the list of all the available regions of interest (ROI's) on the PhenoCam server and producing a data table
myrois <- rois

myrois <- rois %>%
  select(roitype, roi_name, site, active, everything())

myrois
```

### Download time series

Now let's download a time series from a site that you are interested in
exploring. We will select the roi_name, site, and roitype from the
'myrois' data table to make a new data table that contains only these
columns.

[Look at the data table from above and select a site from the site
column that ends in the \_1000 suffix and gather the roitype and
roi_name information for that site.]{style="color:red"} It may be better
to choose a site that is active, where active is TRUE. Once you have
selected your site, [replace the \$roi_name, $site, and$roitype in the
code below with the actual site name, roi type, and roi
name.]{style="color:red"}

Once you have done this, you can now [run]{style="color:green"} the code
chunk below.

```{r}
# Create a new DataFrame with the site information and vegetation type for a site that you may be interested in
selected_columnsdf <- data.frame(roiName = myrois$roi_name, Site = myrois$site, Vegetation_Type = myrois$roitype)
#Downloading the PhenoCam data
selected_columnsdf
```

Great! Now you have selected a site that has an roiName ending in
\_1000.

[For example]{style="color:green"}, you could have picked site 'anisclo'
with the vegetation type == 'EB' and the roiName ending in \_1000. Now
let's use the `get_pheno_ts()` function to download a time series and
return the result as a `data.table` for your site.

[To do this]{style="color:green"}, make sure that you insert the site
name that you selected from the code chunk above from the 'site' column,
replacing the 'insertsitename.' with that site name.

[You will also want to replace 'InsertVegetationType' with the
vegetation_type abbreviation that corresponds to the site you
selected.]{style="color:green"}

This will return a data frame called 'mysite_DB_1000'.
[Run]{style="color:green"} the chunk of code below once the you have
done this.

```{r, fig.height=5, fig.width=6.5, message=FALSE}
# list ROIs for dukehw
myrois[site=='InsertSiteName',]

# Obtain the decidous broadleaf, ROI ID 1000 data from the dukehw phenocam
mysite_DB_1000 <- get_pheno_ts(site = 'InsertSiteName', vegType = 'InsertVegetationType', roiID = 1000, type = '3day')

# Produces a list of the dukehw data variables
str(mysite_DB_1000)

```

Now let's look at the Green Chromatic Coordinate (GCC) from your site.
Remember that this is a measure of "greenness" of an area and is widely
used in Phenocam images as an indicator of the green pigment in
vegetation. Let's use this measure to look at changes in GCC over time
at your site.

We will use `gcc90`, corresponding to the 90th quantile of GCC in the
pixels across the ROI. This tracks the upper greenness values while not
including many outliers.

Before we can plot `gcc-90` we do need to fix our dates and convert them
from Factors to Date to correctly plot. [Run]{style="color:green"} the
chunk of code below without adding any changes.

```{r plot-gcc90, fig.height=5, fig.width=8}
# Convert date variable into date format
mysite_DB_1000[,date:=as.Date(date)]

# plot gcc_90
mysite_DB_1000[,plot(date, gcc_90, col = 'green', type = 'b')]
mtext('My Site GCC', font = 2)

```

### Download midday images for a given time range

Now we can access all the midday images and download them one at a time.
However, we frequently want all the images within a specific time range
of interest. We'll learn how to do that next.

[Replace 'MySiteName' with the name of the site you selected
above.]{style="color:green"}

Feel free to adjust the y = '2017' to a different year or months =
'1:12' if you do not want to include all 12 months, or days = '15' if
you want to choose a different day of the month.

[Once you have made these adjustments, you can run the code chunk
below.]{style="color:green"}

[Note:]{style="color:red"} If y = 2017 does not work, replace this with
y = '2023'.

```{r midday-time-range, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}

# open a temporary directory
tmp_dir <- tempdir()

# download a subset. Example dukehw 2017
download_midday_images(site = 'MySiteName', # which site
                       y = 2017, # which year(s)
                       months = 1:12, # which month(s)
                       days = 15, # which days on month(s)
                       download_dir = tmp_dir) # where on your computer

# list of downloaded files
mysite_middays_path <- dir(tmp_dir, pattern = 'MySiteName', full.names = TRUE)

head(mysite_middays_path)

```

We can demonstrate the seasonality of your site observed from the
camera. ([Note]{style="color:red"} this code may take a while to run
through the loop). All you will need to do for this is
[run]{style="color:green"} the code below without any changes.

```{r plot-monthly-forest, fig.height=6, fig.width=8, message=FALSE, eval=TRUE}
n <- length(mysite_middays_path)
par(mar= c(0,0,0,0), mfrow=c(4,3), oma=c(0,0,3,0))

for(i in 1:n){
  img <- readJPEG(mysite_middays_path[i])
  plot(0:1,0:1, type='n', axes= FALSE, xlab= '', ylab = '')
  rasterImage(img, 0, 0, 1, 1)
  mtext(month.name[i], line = -2)
}
mtext('Seasonal variation from my site', font = 2, outer = TRUE)

```

# Phenology Transition Dates

# Phenology Transition Dates

Now let's explore the GCC, rising, and falling of the seasons on a plot
through time. This let's us visualize how the phenology might have
changed through time in an ecosystem. To do this, we can use the
phenocamr package.

We will use the phenphases() function in the phenocamr package, but
first let's download data frame the site that you have been working with
in this exercise.

[Replace "MySiteName" with the name of your site that you selected. You
will also want to replace "MyVegetationTypeAbbreviation" with your
vegetation type abbreviation that corresponds to your
site.]{style="color:green"}

Now [run]{style="color:green"} the code with these changes.

```{r}
 download_phenocam(site = "MySiteName",
                    veg_type = "MyVegetationTypeAbbreviation",
                    roi_id = "1000",
                    frequency = 3,
                    outlier_detection = FALSE,
                    smooth = FALSE,
                    out_dir = tempdir())

```

You will now want to put the data that you just downloaded into a
data.table.

To do this, you will want to [replace 'MySiteName' and
'MyVegetationAbbreviation' from the file name
"MySiteName_MyVegetationAbbreviation_1000_3day.csv" with your site name
and vegetation abbreviation.]{style="color:green"}

[For example:]{style="color:green"}, if I chose the the site name of
'dukehw' and the vegetation type abbreviation is 'DB' the .csv name
would be 'dukehw_DB_1000_3day.csv'. Once you have made the appropriate
adjustments, you can run the code chunk below. This will output a new
data.table called 'dfphenophases'.

```{r}
#Now we need to assign a dataframe to the data that we just downloaded
dfphenophases <- read_phenocam(file.path(tempdir(),"MySiteName_MyVegetationAbbreviation_1000_3day.csv"))
```

# Detecting Outliers and Smoothing the Time Series

Now that we have our data table for our site, we will want to detect any
outliers in our data and smooth the time series.
[Run]{style="color:green"} the code chunk below without any changes.

```{r}
  dfphenophases <- detect_outliers(dfphenophases)
  dfphenophases <- smooth_ts(dfphenophases)
```

# Calculating the Phenophases

Now that the smoothed data has been made available, we can calculate and
plot the phenological transition dates for your site.
[Run]{style="color:green"} the code chunk below to do this without any
changes to the code!

```{r}
myphenology_dates <- phenophases(dfphenophases, internal = TRUE)

plot(as.Date(dfphenophases$data$date),
     dfphenophases$data$smooth_gcc_90,
     type = "l",
     xlab = "date",
     ylab = "Gcc")

# rising "spring" greenup dates
abline(v = myphenology_dates$rising$transition_50,
       col = "forestgreen")

# falling "autumn" senescence dates
abline(v = myphenology_dates$falling$transition_50,
       col = "brown4")
```

------------------------------------------------------------------------
